Combined annotation exploration for SCPCP000006

### 2025-03-07 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
`r table(cell_type_df$first.level_annotation)["unknown"]`cells remain _unknown_

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation), shape = 19, size = 1) +
  geom_point() +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))

DT::datatable(table(cell_type_df$first.level_annotation, cell_type_df$sample_id))

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "no CNV" ~ "kidney",

    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "no CNV" ~ "normal stroma",

    cell_type_df$compartment %in% c("normal") &
      cell_type_df$cell_type %in% c("endothelial cell") ~ "endothelium",
    cell_type_df$compartment %in% c("normal") &
      cell_type_df$cell_type %in% c("B cell",
                                    "CD4-positive, alpha-beta T cell",
                                    "conventional dendritic cell",
                                    "lymphocyte",
                                    "macrophage",
                                    "mast cell",
                                    "monocyte",
                                    "natural killer cell",
                                    "neutrophil",
                                    "plasmacytoid dendritic cell") ~ "immune",

    # assign cancer cells based on condition
    cell_type_df$compartment %in% c("stroma") &
      cell_type_df$has_cnv_score == "CNV" ~ "cancer stroma",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type == "mesenchymal cell" ~ "blastema",
    cell_type_df$compartment %in% c("fetal_nephron") &
      cell_type_df$has_cnv_score == "CNV" &
      cell_type_df$cell_type != "mesenchymal cell" ~ "cancer epithelium",
    .default = "unknown"
  ))

cell_type_df$second.level_annotation[cell_type_df$sample_id %in% c("SCPCS000177", "SCPCS000180", "SCPCS000181", "SCPCS000190", "SCPCS000197")] <- "unknown"
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q")))
}
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in 1:22) {
  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q")))
}
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in c(1:12, 16:20)) {
  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p")))
}
```

@@

```{r fig.width=20, fig.height=5, out.width='100%', results='asis'}
for (i in c(1:12, 16:20)) {
  print(do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p")))
}
```

@@

CNV_profile_gain <- cell_type_df_sub |>
  select(sample_id, contains("has_dupli_chr")) |>
  group_by(sample_id) |>
  summarise_all(.funs = mean) |>
  pivot_longer(cols = -sample_id) |>
  rename("chromosome_arm" = "name", "percentage" = "value") |>
  left_join( chromosome_arms, by = c("chromosome_arm"="dupli")) |>
  mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
  mutate(
    absolute_70 = ifelse(percentage > 0.7, 1, 0), 
    absolute_10 = ifelse(percentage > 0.1, 1, 0)
 )

# plot

loss_profile <- ggplot(CNV_profile_loss, aes(x = chromosome_arm, y = absolute_10, width = 20 * Size)) +
  geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#41b6c4") +
  scale_x_discrete(labels = gsub("has_loss_", "", levels(droplevels(CNV_profile_loss$chromosome_arm))))  +
  theme_classic() +
  ylab("Number of occurrence") +
  xlab("chromosom arms") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))

gain_profile <- ggplot(CNV_profile_gain, aes(chromosome_arm, y = absolute_10, width = 20* Size)) +
  geom_bar(stat = "identity",  fill = "#ce1256", alpha = 0.4) +
  geom_bar(aes(x = chromosome_arm, y = absolute_70, width = 20 * Size), stat = "identity",  fill = "#ce1256") +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), 
        text = element_text(size = 15)) +
  ylab(" ")

# combined plot
wrap_plots(gain_profile, loss_profile, ncol = 1)

>>>>

Using this basic strategy, we identified `r table(cell_type_df$first.level_annotation)["cancer"]` _cancer_ cells, `r table(cell_type_df$first.level_annotation)["normal"]` _normal_ cells.
`r table(cell_type_df$first.level_annotation)["unknown"]` cells remain _unknown_

```{r fig.width=20, fig.height=20, out.width='100%', results='asis'}
ggplot(cell_type_df, aes(x = umap.umap_1, y = umap.umap_2, color = first.level_annotation)) +
  geom_point(size = 0.5, shape = 19, alpha = 0.5) +
  facet_wrap(facets = ~sample_id, ncol = 5) +
  theme_bw() +
  theme(text = element_text(size = 22))

cell_type_df |>
  dplyr::count(first.level_annotation, sample_id) |>
  tidyr::pivot_wider(
    names_from = first.level_annotation, 
    values_from = n
  )|>
  DT::datatable()

@@

```{r fig.width=10, fig.height=10, out.width='100%', results='asis'}
# Define a vector of `fetal_kidney_predicted.cell_type` levels that corresponds to immune cells
immune_celltypes <- c("B cell",
                                    "CD4-positive, alpha-beta T cell",
                                    "conventional dendritic cell",
                                    "lymphocyte",
                                    "macrophage",
                                    "mast cell",
                                    "monocyte",
                                    "natural killer cell",
                                    "neutrophil",
                                    "plasmacytoid dendritic cell")

cell_type_df <- cell_type_df |>
  mutate(second.level_annotation = case_when(
    # assign normal cells based on condition
      (sample_id                                         %in% none_reference_samples) ~ "unknown",

      (cell_type_df$fetal_kidney_predicted.compartment   == c("fetal_nephron") &
      cell_type_df$cnv_score                             == "no CNV") ~ "kidney",

      (cell_type_df$fetal_kidney_predicted.compartment   == c("stroma") &
      cell_type_df$cnv_score                             == "no CNV") ~ "normal stroma",

      (cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
      cell_type_df$fetal_kidney_predicted.cell_type      == c("endothelial cell")) ~ "endothelium",

      (cell_type_df$fetal_kidney_predicted.compartment   == c("normal") &
      cell_type_df$fetal_kidney_predicted.cell_type    %in% immune_celltypes) ~ "immune",

    # assign cancer cells based on condition
      (cell_type_df$fetal_kidney_predicted.compartment    == c("stroma") &
      cell_type_df$cnv_score                              == "CNV") ~ "cancer stroma",

      (cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
      cell_type_df$cnv_score                              == "CNV" &
      cell_type_df$fetal_kidney_predicted.cell_type       == "mesenchymal cell") ~ "blastema",

      (cell_type_df$fetal_kidney_predicted.compartment    == c("fetal_nephron") &
      cell_type_df$cnv_score                              == "CNV" &
      cell_type_df$fetal_kidney_predicted.cell_type       != "mesenchymal cell") ~ "cancer epithelium",

@@

```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
# create a list of all plots by mapping over the vector of indices, instead of a for loop
1:22 |>
  purrr::map(
    \(i) 
      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "q"))
) |>
# send into patchwork to make a single plot from the list of plots in 2 columns
patchwork::wrap_plots(ncol = 2)
```

@@

```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
# create a list of all plots by mapping over the vector of indices, instead of a for loop
1:22 |>
  purrr::map(
    \(i) 
      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "q"))
) |>
# send into patchwork to make a single plot from the list of plots in 2 columns
patchwork::wrap_plots(ncol = 2)
```

@@

```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
# create a list of all plots by mapping over the vector of indices, instead of a for loop
c(1:12, 16:20) |>
  purrr::map(
    \(i) 
      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_loss_chr", i, "p"))
) |>
# send into patchwork to make a single plot from the list of plots in 2 columns
patchwork::wrap_plots(ncol = 2)
```

@@

```{r fig.width=25, fig.height=50, out.width='100%', results='asis'}
# create a list of all plots by mapping over the vector of indices, instead of a for loop
c(1:12, 16:20) |>
  purrr::map(
    \(i) 
      do_Feature_mean(cell_type_df_sub, group.by = "second.level_annotation", feature = glue::glue("has_dupli_chr", i, "p"))
) |>
# send into patchwork to make a single plot from the list of plots in 2 columns
patchwork::wrap_plots(ncol = 2)
```

@@

CNV_profile_gain <- cell_type_df_sub |>
                    select(sample_id, contains("has_dupli_chr")) |>
                    group_by(sample_id) |>
                    summarise_all(.funs = mean) |>
                    pivot_longer(
                      cols = -sample_id, 
                      names_to = "chromosome_arm", 
                      values_to = "percentage") |>
                    left_join(chromosome_arms, by = c("chromosome_arm"="dupli")) |>
                    mutate(chromosome_arm = factor(chromosome_arm, levels = has_dupli_chr)) |>
                    mutate(
                      absolute_70 = ifelse(percentage > 0.7, 1, 0), 
                      absolute_10 = ifelse(percentage > 0.1, 1, 0))

# plot
# Here we chose to loop on chromosome arms to then be able to scale each bar to the length of the corresponding genomic region using wrap_plot. 
# Of note, the resulting plot is way better than the one obtained with facet_wrap or facet_grid so far.

loss_profile <- list()
for(chr in unique(CNV_profile_loss$chromosome_arm)){
  loss_profile[[chr]] <- ggplot(CNV_profile_loss[CNV_profile_loss$chromosome_arm == chr,], 
                                aes(x = chromosome_arm, y = absolute_10)) +
                        geom_bar(stat = "identity",  fill = "#41b6c4", alpha = 0.4) +
                        geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#41b6c4") +
                        scale_x_discrete(labels = gsub("has_loss_", "", chr))  +
                        theme_void() +
                        ylim(c(-40,0)) +
                        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12), text = element_text(size = 15))
}
loss_profile <- wrap_plots(loss_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$loss %in% unique(CNV_profile_loss$chromosome_arm)], guides = "collect")

gain_profile <- list()
for(chr in unique(CNV_profile_gain$chromosome_arm)){
  gain_profile[[chr]] <- ggplot(CNV_profile_gain[CNV_profile_gain$chromosome_arm == chr,], aes(x = chromosome_arm, y = absolute_10)) +
  geom_bar(stat = "identity",  fill = "#ce1256", alpha = 0.4) +
  geom_bar(aes(x = chromosome_arm, y = absolute_70), stat = "identity",  fill = "#ce1256") +
  theme_void() +
  ylim(c(0,40)) +
  theme(axis.text.x = element_blank())
}
gain_profile <- wrap_plots(gain_profile, nrow = 1, widths = chromosome_arms$Size[chromosome_arms$dupli %in% unique(CNV_profile_gain$chromosome_arm)], guides = "collect")

# combined plot
wrap_plots(gain_profile, loss_profile, ncol = 1) + plot_layout( widths = CNV_profile_gain$Size)
